### 2025-06-05 paper/theory.tex
Update results.

\subsubsection{Advantage-Weighted Transformer Training} A key innovation in POLARIS is the use of advantage-weighted training for the belief processor:
\begin{align}
    J^i_{\text{transformer}}(\psi^i_{\text{Transformer}}) &= \mathbb{E}_{D^i}\left[A^i(b^i, a^i) \cdot \text{BCE}(\mathbb{P}(s|b^i), o^i_{t+1})\right]
\end{align}

where $A^i(b^i, a^i)$ is the advantage function and BCE is binary cross-entropy loss. This objective bridges belief formation and policy optimization in a novel way. By weighting the belief update loss with the advantage function, the Transformer is encouraged to generate belief states that not only accurately predict future observations but also lead to high-value actions. In traditional partially observable reinforcement learning, belief updates are typically performed independently of policy optimization. By contrast, POLARIS creates a direct pathway for policy performance to influence belief formation, prioritizing informative aspects of the state space that have the greatest impact on decision quality. This approach is particularly valuable in social learning contexts, where certain beliefs about others' strategies may be more consequential than others for effective coordination or competition.

\subsubsection{Catastrophic Forgetting Prevention}

In social learning environments, the underlying state of the world often remains fixed during an episode, but varies across episodes. This creates a continual learning challenge: agents must adapt to new states while retaining knowledge about previously encountered states. Without proper mechanisms, neural networks tend to overwrite previously learned representations when adapting to new conditions—a phenomenon known as catastrophic forgetting \citep{vandeven2024continuallearningcatastrophicforgetting}. POLARIS addresses this challenge through Elastic Weight Consolidation (EWC), which selectively preserves important parameters for previously encountered states while allowing flexibility for adaptation. EWC augments the standard training objective with a regularization term that penalizes significant changes to parameters critical for performance in previously encountered states:

\begin{align}
    L_{\text{EWC}}(\theta) = L(\theta) + \sum_i \frac{\lambda}{2} F^i (\theta^i - {\theta^i}^*)^2
\end{align}

where $\theta^*$ represents parameters that performed well on previous tasks, $F^i$ is the Fisher information matrix that measures parameter importance, and $\lambda$ controls the strength of regularization. The Fisher information matrix provides a principled way to identify which parameters are most crucial for previously learned tasks, calculated using gradients of the log-likelihood of the model's output with respect to the parameters on data from previously encountered states.

Crucially, POLARIS introduces a dynamic mechanism that automatically adjusts the EWC importance factor based on observed performance:

\begin{align}
    \lambda_t = 
    \begin{cases}
        \lambda_{t-1} \cdot 0.8 & \text{if loss ratio} > 1.5 \text{ (loss increasing too much)} \\
        \lambda_{t-1} \cdot 1.2 & \text{if loss ratio} < 0.5 \text{ (loss decreasing too quickly)} \\
        \lambda_{t-1} & \text{otherwise}
    \end{cases}
\end{align}

This adaptive approach monitors the ratio between current and previous loss values, reducing the EWC importance when the loss is increasing too rapidly (indicating difficulty learning new information due to excessive preservation of old parameters) and increasing it when the loss is decreasing too quickly (suggesting potential forgetting of valuable past knowledge). The implementation also includes both L1 and L2 penalties with adaptive weighting between them, transitioning from more L2 regularization for early tasks to more L1 regularization for later tasks, encouraging sparser updates as the model accumulates more knowledge from different states.

@@

\begin{algorithm}[H]
\caption{POLARIS Training Algorithm for Single Agent}
\begin{algorithmic}[1]
\State Initialize belief processor, inference module, policy and value networks with parameters $\psi^i_{\text{Transformer}}$, $\phi^i$, $\theta^i$, $\psi^i_{\beta}$
\State Initialize replay buffer $D^i$, initial belief $b^i_0$, and latent state $\boldsymbol{\hat{z}}^{-i}_0$
\State Determine action space type (discrete or continuous)
\For{each step $t$}
    \State Observe signal $o^i_t$ and joint action $\boldsymbol{a}_{t-1}$
    \State Update belief: $b^i_t, \mathbb{P}(s|b^i_t) = f_{\text{Transformer}}(b^i_{t-1}, o^i_t, \boldsymbol{a}_{t-1}; \psi^i_{\text{Transformer}})$
    \State Select action: 
    \If{discrete action space}
        \State $a^i_t \sim \pi^i(\cdot|b^i_t, \boldsymbol{\hat{z}}^{-i}_t; \theta^i)$ \Comment{Sample from categorical distribution}
    \Else
        \State $\mu, \sigma = \text{MLP}_{\text{policy}}(b^i_t, \boldsymbol{\hat{z}}^{-i}_t; \theta^i)$
        \State $\epsilon \sim \mathcal{N}(0, I)$
        \State $a^i_t = \tanh(\mu + \sigma \odot \epsilon)$ \Comment{Sample using reparameterization trick}
    \EndIf
    \State Execute action, observe reward $r^i_t$ and next signal $o^i_{t+1}$
    \State Infer latent: $\boldsymbol{\hat{z}}^{-i}_{t+1} = f_{\text{GNN}}(o^i_t, \boldsymbol{a}_t, r^i_t, o^i_{t+1}; \phi^i)$
    \State Store transition in replay buffer
    \If{update step}
        \State Sample batch $B$ from replay buffer $D^i$
        \State Update inference module: $\phi^i \leftarrow \phi^i - \alpha_{\phi} \nabla_{\phi^i} J^i_{\text{elbo}}(B)$
        \State Update value networks: $\psi^i_{\beta} \leftarrow \psi^i_{\beta} - \alpha_{\psi} \nabla_{\psi^i_{\beta}} J^i_{q}(B)$
        \State Update policy: $\theta^i \leftarrow \theta^i - \alpha_{\theta} \nabla_{\theta^i} J^i_{\pi}(B)$ \Comment{Using appropriate entropy calculation}
        \State Update belief processor: $\psi^i_{\text{Transformer}} \leftarrow \psi^i_{\text{Transformer}} - \alpha_{\psi} \nabla_{\psi^i_{\text{Transformer}}} J^i_{\text{transformer}}(B)$
        \State Update target networks: $\bar{\psi}^i_{\beta} \leftarrow \tau \psi^i_{\beta} + (1-\tau) \bar{\psi}^i_{\beta}$
        \If{true state has changed}
            \State Calculate Fisher matrices from current replay buffer
            \State Register task with EWC
        \EndIf
    \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

The algorithm follows a sequential process where each agent first updates its belief state based on new observations and the joint actions of all agents using the Transformer-based belief processor (line 6). Based on this updated belief and the current inference of other agents' latent states, the agent selects an action using its stochastic policy—sampling from a categorical distribution for discrete action spaces (line 9) or using the reparameterization trick for continuous action spaces (lines 11-13). After executing this action and observing the reward and next observation (line 15), the agent updates its inference of other agents' latent states using the GNN-based inference module (line 16). All of these experiences are stored in a replay buffer for later training (line 17).

At designated update steps, the agent samples a batch of transitions from its replay buffer and uses these to update all network components. The inference module is updated first (line 19) to improve predictions of other agents' actions using the ELBO objective. Next, the value networks are updated (line 20) to better estimate expected returns. The policy network is then updated (line 21) to maximize expected value and entropy. Finally, the belief processor is updated (line 22) using the advantage-weighted training approach. Target networks, used for stable learning in the SAC framework, are updated using a weighted average of the current and target parameters (line 23).

A crucial feature of POLARIS is its adaptive response to changes in the environment state. When the true state changes (line 24), the algorithm calculates Fisher information matrices from the current replay buffer (line 25) and registers a new task with the EWC mechanism (line 26), preserving important knowledge from the previous state while allowing adaptation to the new state.

>>>>

@@

\begin{algorithm}[!htbp]
\caption{POLARIS Training Algorithm for Single Agent}
\begin{algorithmic}[1]
\State Initialize belief processor, inference module, policy and value networks with parameters $\psi^i_{\text{Transformer}}$, $\phi^i$, $\theta^i$, $\psi^i_{\beta}$
\State Initialize replay buffer $D^i$, initial belief $b^i_0$, and latent state $\boldsymbol{\hat{z}}^{-i}_0$
\State Determine action space type (discrete or continuous)
\For{each step $t$}
    \State Observe signal $o^i_t$ and joint action $\boldsymbol{a}_{t-1}$
    \State Update belief: $b^i_t, \mathbb{P}(s|b^i_t) = f_{\text{Transformer}}(b^i_{t-1}, o^i_t, \boldsymbol{a}_{t-1}; \psi^i_{\text{Transformer}})$
    \State Select action: 
    \If{discrete action space}
        \State $a^i_t \sim \pi^i(\cdot|b^i_t, \boldsymbol{\hat{z}}^{-i}_t; \theta^i)$ \Comment{Sample from categorical distribution}
    \Else
        \State $\mu, \sigma = \text{MLP}_{\text{policy}}(b^i_t, \boldsymbol{\hat{z}}^{-i}_t; \theta^i)$
        \State $\epsilon \sim \mathcal{N}(0, I)$
        \State $a^i_t = \tanh(\mu + \sigma \odot \epsilon)$ \Comment{Sample using reparameterization trick}
    \EndIf
    \State Execute action, observe reward $r^i_t$ and next signal $o^i_{t+1}$
    \State Infer latent: $\boldsymbol{\hat{z}}^{-i}_{t+1} = f_{\text{GNN}}(o^i_t, \boldsymbol{a}_t, r^i_t, o^i_{t+1}; \phi^i)$
    \State Store transition in replay buffer
    \If{update step}
        \State Sample batch $B$ from replay buffer $D^i$
        \State Update inference module: $\phi^i \leftarrow \phi^i - \alpha_{\phi} \nabla_{\phi^i} J^i_{\text{elbo}}(B)$
        \State Update value networks: $\psi^i_{\beta} \leftarrow \psi^i_{\beta} - \alpha_{\psi} \nabla_{\psi^i_{\beta}} J^i_{q}(B)$
        \State Update policy: $\theta^i \leftarrow \theta^i - \alpha_{\theta} \nabla_{\theta^i} J^i_{\pi}(B)$ \Comment{Using appropriate entropy calculation}
        \State Update belief processor: $\psi^i_{\text{Transformer}} \leftarrow \psi^i_{\text{Transformer}} - \alpha_{\psi} \nabla_{\psi^i_{\text{Transformer}}} J^i_{\text{transformer}}(B)$
        \State Update target networks: $\bar{\psi}^i_{\beta} \leftarrow \tau \psi^i_{\beta} + (1-\tau) \bar{\psi}^i_{\beta}$
        \If{true state has changed}
            \State Calculate Fisher matrices from current replay buffer
            \State Register task with EWC
        \EndIf
    \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

The algorithm follows a sequential process where each agent first updates its belief state based on new observations and the joint actions of all agents using the Transformer-based belief processor. Based on this updated belief and the current inference of other agents' latent states, the agent selects an action using its stochastic policy—sampling from a categorical distribution for discrete action spaces or using the reparameterization trick for continuous action spaces. After executing this action and observing the reward and next observation, the agent updates its inference of other agents' latent states using the GNN-based inference module. All of these experiences are stored in a replay buffer for later training.

At designated update steps, the agent samples a batch of transitions from its replay buffer and uses these to update all network components. The inference module is updated first to improve predictions of other agents' actions using the ELBO objective. Next, the value networks are updated to better estimate expected returns. The policy network is then updated to maximize expected value and entropy. Finally, the belief processor is updated using the advantage-weighted training approach. Target networks, used for stable learning in the SAC framework, are updated using a weighted average of the current and target parameters.

A crucial feature of POLARIS is its adaptive response to changes in the environment state. When the true state changes, the algorithm calculates Fisher information matrices from the current replay buffer and registers a new task with the EWC mechanism, preserving important knowledge from the previous state while allowing adaptation to the new state.
